Diversity and evolution of centromere repeats in the maize genome

### 2014-05-02 centc.tex
merge test

We used principle coordinate analysis (PCoA) to cluster CENTC variants based on their genetic distances. Eigenvalues from the PCoA were used to determine the number of statistically significant clusters using the Tracy-Widom distribution \citep{Patterson2006}.

We employed the software SpaGeDi (\citep{Hardy2002}\url{http://ebe.ulb.ac.be/ebe/Software.html}) to estimate the spatial autocorrelation of sequence similarity of CENTC repeats in the completely sequenced centromeres 2 and 5.  We calculated Moran’s I statistic using Jukes-Cantor genetic distance and measures of physical distance between CENTC repeats in base pairs.  Confidence intervals for the values of I were estimated by 20,000 random permutations of the physical distances.

@@

%We wanted to better understand the diversity present in all CentC copies across the genome.  Looking at our alignment, we found that the sequences had a pairwise percent identify of 65.8\% with only 2.2\% of sites identical, suggesting large sequence diversity.  However, these statistics do not inform us whether the amount of diversity observed would be expected due to cumulative mutation over time.  
In order to better understand patterns of diversity at CentC, we performed simulations to test the likelihood of homoplasious mutations (i.e. independent mutations occurring at the same position in two different CentC repeats). Our simulation (code available at: \jri{need this github URL!}) assumed that CentC has been evolving for  1 million years since the divergence of maize and \emph{Tripsacum} \citep{Ross-Ibarra2009}, a closely related genus whose centromere repeat shares a large amount of homology \citep{Melters2012}.  We assumed a constant copy number, a mutation rate of $3 x 10^{-8}$ (\jri{CITE Clark 2005 MBE}), and one generation per year.

@@

We found a total of 12,162 CentC variants in the maize reference genome and  unassembled BACs.  Of these, 8,259 were unique over their full length. No CentC sequence occurred more than 10 times in the genome, and the vast majority ($>$75\%, Supplemental Table \ref{supp.occurence}) of non-unique CentC variants occurred only twice.  Of the 2,266 non-unique CentC sequences, only 3 were tandem, identical duplicates.  Genome-wide CentC locations also show that nearly all of the 10,639 \jri{why is this different from 8259 number?} CentC copies are found in one of the 248 clusters identified on chromosomes 1-10; only 14 occurred as solo copies.  Clusters varied in width from single CentC copies to 84KB \jri{check is it KB or Kb?}, with a mean of $\sim$7KB ($\sim$45 CentC copies). Chromosomes varied greatly in CentC copy number, though we know that centromere assemblies for all of the chromosomes are not complete.   For example, CENH3-ChIP sequence from an oat-maize addition line including only one maize chromosome \jri{CITE} nonetheless has many reads that map to the unassembled BACs (Supplemental table \ref{gaby}).  In particular, chromosome 6 had many more reads aligning to the unassembled BACs than it did to its own centromere, suggesting a particularly incomplete assembly.  Examining total repeat number, Chromosome 7 had the most CentC, with 3,200 copies, while chromosome 6 had the fewest with 32 copies.  \jri{consistency check: is Chromosome capitalize?}

@@

We then identified how many large clusters of CentC were retained from each of the two parental genomes that comprise the extant maize genome, referred to here as subgenome 1 and subgenome 2 (Figure \ref{circos}).  Previous work identified the parental genome for individual chromosomal segments \citep{Schnable2011} and centromeres \citep{Wang2012}.  Because large clusters are less likely to be misassembled, we focused our analyses on the 52 clusters $>$10KB in length (Fig \ref{Supp_clusterhist}).  We assign clusters to a subgenome if they are flanked by two regions identified as originating from the same  subgenome.  Thirty-eight of these clusters could be assigned to subgenome 1 (out of 43 assignable). If we restrict the analysis to clusters  $>$20KB, all of the 16 unambiguously assignable clusters come from subgenome 1.  Even correcting for the genome-wide overrepresentation of subgenome 1 (62.5\% of assigned base pairs), these results suggest a strong inequality in the origin of large CentC clusters (Fisher's exact test, $p<0.005$ for both 10KB and 20KB clusters).

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
%  \includegraphics[width=0.75\textwidth]{circos.png}
  \includegraphics{circos.png}
\caption{CentC repeat location in relation to the maize subgenomes.  The outer ring depicts chromosomal assignment to the two subgenomes, with higher confidence regions are colored with darker colors.  Breakpoints between the subgenomes remain uncolored to indicate uncertainty. The middle ring, shaded in blue, displays the locations of all CentCs across the genome.  The inner ring, shaded in yellow, displays the locations of all CentC clusters greater than 20KB in length.}
\label{circos}    
\end{figure}

@@

CentC copies in the maize genome exhibit tremendous diversity: the overall pairwise identity in our alignment was only 65\%, and $\sim$98\% of sites in the alignment had at least 2 variants.  Such diversity led us to ask whether genetic groups of CentC variants could be distinguished. We performed principle coordinate  \jri{principal coordinate or principal component?} analyses (PCoA) from a genetic distance matrix estimated from our alignment, and assigned individual repeats to  genetic clusters following the approach of \citet{Patterson2006}.  We found 58 significant clusters, but observed no pattern of groupings that revealed chromosome specificity of CentC’s or the impact of historical tetraploidy (Figure \ref{pcoa}; Supplemental Table \ref{long}).

\begin{figure}[h]
\centering
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\includegraphics[width=1\textwidth]{Rplot02}
\caption{Presence of CentC in each of the heirarchical groups.  The 58 clusters found to be statistically significant in forming genetic groups are represented on the x-axis and chromosome of origin on the y-axis. The size of each point is proportional to the log number of sequences in that group on that chromosome. CentC counts from chromosomes whose centromeres were derived from subgenome 1 are colored green and those from subgenome 2 are colored brown. 
}
\label{pcoa}    
\end{figure}

@@

\begin{figure}[h]
\centering
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\includegraphics[width=1\textwidth]{heatmap.png}
\caption{CentC physical location and genetic relatedness for (a) chromosome 2 and (b) chromosome 5. \jri{need a and b labels } On the physical map, red lines show locations of numbered CentC clusters and blue blocks show the location of the active kinetichores.  Scale bar is in MB.  Below each physical map is shown a heatmap of genetic relatedness of each CentC to (top row) other copies within its island of tandem repeats delineated by dotted lines and (bottom row) all other copies on the chromosome.  Darker colors indicate higher relatedness.  The total number of CentC in each cluster is shown below the map.  
}
\label{heatmap}    
\end{figure}

The decreased genetic distance among CentCs in local clusters on chromosome 2 and 5 suggest that many of the genetic groupings discovered in our genome-wide analysis should correspond to local clusters of repeats. We see little evidence of this, however, as repeats within individual clusters are frequently found in different genetic groups as defined by principle component \jri{component or coordinate?} analysis (Figure \ref{pcoa}).  A comparison of all pairs of CentC sequences reveals a likely explanation: of the $\sim 74$ million possible pairs,  approximately 6 million share $\geq 2$ mutations different from the genome-wide consensus, likely grouping in genetic clusters despite their physical distance.  Looking at several triplets at random from our alignment confirms that two sequences in one PCoA assignment sometimes share greater pairwise identity than two sequences adjacent to one another in different PCoA groups.  A simple forward simulation (see Methods) suggests this pattern could be due entirely to homoplasy rather than long-distance movement of CentC repeats.  By stochastically applying mutations to an initially homogeneous group of repeat sequences, we find that plausible parameter values produce $\sim 10$ million pairs of repeats sharing $\geq 2$ mutations.

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\centering
\includegraphics[width=1\textwidth]{Figure3_noedge}
\caption{(a) Mb of CentC in genomic libraries of maize and teosinte.  Box plots show data from \citet{Chia2012}.  Points show data for maize inbred B73 and the teosinte \emph{Z. luxurians} from \citet{Tenaillon2011}.  For comparison, the data point of maize inbred B73 in \citet{Chia2012} is shown with a tick mark on the box plot. (b) FISH images from an F1 hybrids between maize and the teosinte \emph{parviglumis}, and maize and the teosinte \emph{Z. luxurians}.  Upper row shows the karyotype for the teosinte and the lower row shows the maize karyotype.  Red=CentC, Green=CRM. 
}
\label{abundance}    
\end{figure}

@@

%While genome size (Poggio et al 1998), transposable elements (TEs) (Chia et al 2012), and heterochromatic knob content (Poggio et al 1998) are known to have changed greatly through domestication, little information exists as to how domestication has impacted centromeric repeats.  Previous cytogenetic work has identified considerable variation in centromere repeat content  between domesticated maize and its wild relatives Z. mays ssp. parviglumis,), Z. mays ssp. mexicana,, and Z. luxurians (hereafter parviglumis, mexicana, and luxurians) (Albert et al 2010).

@@

Our study traces the changes in centromere repeat genetic relatedness across maize chromosomes to show how ancient tetraploidy and subsequent evolution has impacted centromeres.  Most interestingly, we show that most large arrays of CentC in the maize reference genome derive from maize1 subgenome, which is known to have lower gene loss and higher average expression than maize subgenome 2 \cite{Schnable2011}.  Regulatory microRNA’s are known to correspond to centromeric repeats \citep{ReinhartBartel2002}.  Therefore, it may be worthwhile to further investigate the potential correlation between centromere retention and gene expression as the phenomenon of higher expression of genes from one ancestor may be very common across plants (\emph{A. thaliana} \cite{Cheng2012} and \emph{Gossypium} Kapp and Wendel unpublished).  \pb{Two comments about this: first, its a terrible sentence so help make this better, because i think the point is really nice.  second, the best citations i found for cotton was an unpublished ref that is still not published from what i can gather.  advice? e.g. \url{http://www.plosone.org/article/info\%3Adoi\%2F10.1371\%2Fjournal.pone.0036442} also check out papers that cite Schabs: \url{http://scholar.google.com/scholar?cites=5313714612108479532&as_sdt=2005&sciodt=0,5&hl=en} } \jri{see journal policies, but citing unpublished usually kinda lame.} There are several possible explanations for the differential contribution of subgenome 1 to the diversity of CentC in the extant maize  genome. We put forth several hypotheses that may explain the strong bias for CentC to be located within a block from subgenome 1. Subgenome 2 may have had severely reduced quantities of CentC and therefore never contributed equally.  The CentC from subgenome 2 may have decayed rapidly and thus less of it remains.  Lastly, the CentC from subgenome 1 may have expanded in copy number since the allopolyploidy event.  Though we are unable to conclusively identify the reason, the lack of a large number of identical tandem duplicates  and lack of close proximity highest genetic relatedness pairs suggests that any large scale expansion of CentC has not been recent.  Also, CentC’s in subgenome 2 cluster do not appear to have accumulated excess mutation, suggesting that they are not decaying more rapidly than their subgenome 1 counterparts.  We therefore believe that it is most likely that the subgenomes had an unequal contribution of CentC at the formation of the allopolyploid, though more research about the ancient parents is required.

We also explored how genetic relatedness between CentC’s correlated with location along the physical map.  According to our PCoA analyses, CentC copies across the genome do not form distinct genetic groups correlating with past origin (Figure \ref{pcoa}).  The lack of subgenome or chromosome grouping in modern CentC suggests a little differentiation between the repeats since the ancestral divergence.  Given the lack of a genome wide pattern, we also wanted to investigate genetic relatedness on individual chromosomes.  We hypothesized that most copies of CentC arose from local duplication rather than transposition, and therefore the genetic distance between CentC’s would be lowest across CentC’s within a large tandem array.  \pb{This part is ugly, and needs help} From the reference genome, we chose to investigate chromosomes 2 and 5, since they have been sequenced from end to end, allowing for inferences about the relationship between diversity and physical proximity \citep{Wolfgruber2009}.  Furthermore, the role of the repeat arrays on chromosomes 2 and 5 appear very different, as the largest array on chromosome 2 interacts with the kinetochore, while no array on chromosome 5 does \citep{Wolfgruber2009}.  We find that repeats on chromosomes 2 and 5 are most highly genetically related to neighboring copies (Figure \ref{heatmap}).  This relationship was recapitulated in analyses using SpaGeDi, where CentC’s within approximately 10-50KB of one another were more genetically similar (Supplemental Data \ref{Supp_SpagediChr02} and \ref{Supp_SpagediChr05}).  Though we did not investigate the relationship between CentC location and genetic similarity on the other chromosomes due to incomplete sequencing of centromeres, we observe that many CentC’s within an array fall into the same  significant grouping in our Tracy-Widom clustering analysis, suggesting that high local similarity of CentC’s is a genome-wide phenomenon.  This local relatedness suggests that most CentC is evolving through tandem duplication and not a long distance mechanism such as transposition that had been previously suggested \citep{Shi2010}.  Concerning chromosomes 2 and 5, the higher local relatedness of all CentC clusters regardless of presence within the kinetochore suggests that CentC interaction with kinetochore proteins does not have a detectable change on their rates of evolution.

@@

When studying CentC changes through domestication, our findings show that the repeat has experienced a decrease in copy number over time without a noticeable change in the structure of the repeat arrays. Using PacBio long read sequencing, we confirm genome-wide the observation we see on chromosomes 2 and 5, that most copies of CentC exist in large tandem arrays in all \emph{Zea} taxa.  From short read sequencing data, we show that modern maize has reduced genomic abundance of CentC when compared to the teosintes, a finding confirmed through FISH (Figure \ref{abundance}).  Both teosintes within the species \emph{mays} and its sister species \emph{luxurians} have elevated levels of CentC.  Lower levels of centromere repeats in inbred maize contrasts previous studies that characterized the changing abundance of other common repeats.  For example, the abundance of most transposable element families increased after domestication \citep{Chia2012}.  Knowing that TE abundance increases in domesticated maize, we might have expected CentC content to increase as well if centromere size had to expand alongside most other repetitive content.  Alternatively, due to their structural role in kinetichore formation, we might have also expected centromere repeats to be largely excluded from the genome wide fluctuations in repetitive content assuming that selection exists to maintain centromere size.  Instead, the decrease in CentC content may indicate an active process of removing CentC.  We hypothesize that the removal of CentC content may correlate with genome size, as the sharp decrease in knob content through domestication actually led to an overall smaller genome size in inbred maize. The correlation between  functional  centromere size and genome size has been observed in grass species \citep{Zhang2012}.  Further investigation would be required to discover whether only those copies of CentC outside of the active kinetochore are being deleted.

We also sought to further characterize the chromosome specific repeat cent4 \citep{Page2001}.  Previous work showed that Cent4 probes lagged behind CentC probes in cell division \citep{Jiang2002}, suggesting that the repeat is not involved in the active kinetochore and rather located in the pericentromere.  A lack of enrichment of cent4 in ChIP data agrees with a pericentromeric location.  Using the published cent4 sequences, we identify the cent4 repeat as the LTR retrotransposon RLX\_sela in the transposable element database \citep{Baucom2009}.  Characterization of the TE consensus sequence shows that it contains repetitive motifs with homology to both knob and telomere repeats.  It lacks any of the protein sequences necessary for autonomous transposition, such as GAG and POL complexes.  Previous work in rice has documented the presence of nonautonomous LTR retrotransposons  in or near the centromere \citep{Jiang2002}.  However, the RLX\_sela sequence also appears to be missing some of the necessary primer binding sites that would distinguish it as a nonautonomous TE, suggesting that it may be a tandem repeat, and further sequencing of the region will be required to uncover the role of cent4.   Given that the centromere and pericentromere on chromosome 4 show signs of strong selection during domestication \citep{Hufford2012}, it is plausible that the increase in abundance of Cent4 seen in domesticated maize is due to the effects of linked selection on a locus important for domestication rather than novel insertions or duplications.

@@

\begin{acknowledgements}
We wish to thank Pacific Biosciences for sequencing resources.  We thank Siddharth Bhadra-Lobo, Vince Buffalo, Anne Lorant, Gernot Presting, Lauren Sagara, Michelle Stitzer, and NSF summer exchange program interns Cesar Alvarez Mejia, Aurelio Hernandez Bautista for advice and helpful discussion. This project was funded by NSF grant IOS-0922703.
\end{acknowledgements}

@@

% For tables use
\begin{table}[]
\caption{ChIP Reads mapping to Unassembled from different Oat-Maize Addition (OMA) Lines}
\label{supp.gaby}       % Give a unique label
% For LaTeX tables use
\begin{tabular}{llll}
\hline\noalign{\smallskip}
File Key & Maize Chr & Percent Reads Aligning to Unassembled BACs\\
\noalign{\smallskip}\hline\noalign{\smallskip}				
JJ1BU (OMA 6.34) & 6 & 21.35\\
JJ1BR (OMA 1.36) & 1 & 15.56\\
JJ1CF (OMA 9.41) & 9 & 6.16\\
JJ1CH (OMA 8.05) & 8 & 6.69\\
JJ1CG (OMA 10.26) & 10 & 9.21\\
JJ1CI (OMA 8.05) & 8 & 8.5\\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

>>>>

We used principle coordinate analysis (PCoA) to cluster CentC variants based on their genetic distances. Eigenvalues from the PCoA were used to determine the number of statistically significant clusters using the Tracy-Widom distribution \citep{Patterson2006}.

We employed the software SpaGeDi (\citep{Hardy2002}\url{http://ebe.ulb.ac.be/ebe/Software.html}) to estimate the spatial autocorrelation of sequence similarity of CentC repeats in the completely sequenced centromeres 2 and 5.  We calculated Moran’s I statistic using Jukes-Cantor genetic distance and measures of physical distance between CentC repeats in base pairs.  Confidence intervals for the values of I were estimated by 20,000 random permutations of the physical distances.

@@

%We wanted to better understand the diversity present in all CentC copies across the genome.  Looking at our alignment, we found that the sequences had a pairwise percent identify of 65.8\% with only 2.2\% of sites identical, suggesting large sequence diversity.  However, these statistics do not inform us whether the amount of diversity observed would be expected due to cumulative mutation over time.  
In order to better understand patterns of diversity at CentC, we performed simulations to test the likelihood of homoplasious mutations (i.e. independent mutations occurring at the same position in two different CentC repeats). Our simulation (code available at: \url{https://github.com/paulbilinski/CentC_Analyses/tree/master/Diversity_sims}) assumed that CentC has been evolving for  1 million years since the divergence of maize and \emph{Tripsacum} \citep{Ross-Ibarra2009}, a closely related genus whose centromere repeat shares a large amount of homology \citep{Melters2012}.  We assumed a constant copy number, a mutation rate of $3 x 10^{-8}$ \citep{clark2005}, and one generation per year.

@@

We found a total of 12,162 CentC variants in the maize reference genome and unassembled BACs.  Of these, 8,259 were unique over their full length. No CentC sequence occurred more than 10 times in the genome, and the vast majority ($>$75\%, Supplemental Table \ref{supp.occurence}) of non-unique CentC variants occurred only twice.  Of the 2,266 non-unique CentC sequences, only 3 were tandem, identical duplicates.  Excluding the 1,523 copies that occur on the unassembled BACs, genome-wide CentC locations also show that nearly all of the 10,639 CentC copies are found in one of the 248 clusters identified on chromosomes 1-10; only 14 occurred as solo copies.  Clusters varied in width from single CentC copies to 84KB with a mean of $\sim$7KB ($\sim$45 CentC copies). Chromosomes varied greatly in CentC copy number, though we know that centromere assemblies for all of the chromosomes are not complete.   For example, CENH3-ChIP sequence from an oat-maize addition line including only one maize chromosome \citep{kynast2001} nonetheless has many reads that map to the unassembled BACs (Supplemental table \ref{supp.gaby}).  In particular, chromosome 6 had many more reads aligning to the unassembled BACs than it did to its own centromere, suggesting a particularly incomplete assembly.  Examining total repeat number, chromosome 7 had the most CentC, with 3,200 copies, while chromosome 6 had the fewest with 32 copies.

@@

We then identified how many large clusters of CentC were retained from each of the two parental genomes that comprise the extant maize genome, referred to here as subgenome 1 and subgenome 2 (Figure \ref{circos}).  Previous work identified the parental genome for individual chromosomal segments \citep{Schnable2011} and centromeres \citep{Wang2012}.  Because large clusters are less likely to be misassembled, we focused our analyses on the 52 clusters $>$10KB in length (Supplementary Fig \ref{Supp_clusterhist}).  We assign clusters to a subgenome if they are flanked by two regions identified as originating from the same  subgenome.  Thirty-eight of these clusters could be assigned to subgenome 1 (out of 43 assignable). If we restrict the analysis to clusters  $>$20KB with clear assignment to one subgenome, all 16 clusters were found in subgenome 1 regions.  Even correcting for the genome-wide overrepresentation of subgenome 1 (62.5\% of assigned base pairs), these results suggest a strong inequality in the origin of large CentC clusters (Fisher's exact test, $p<0.005$ for both 10KB and 20KB clusters).  One cluster $>$20KB falls within an unassigned region on chromosome 3.

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
%  \includegraphics[width=0.75\textwidth]{circos.png}
\includegraphics{Fig1_circos.png}
\caption{CentC repeat location in relation to the maize subgenomes.  The outer ring depicts chromosomal assignment to the two subgenomes, with higher confidence regions are colored with darker colors.  Breakpoints between the subgenomes remain uncolored to indicate uncertainty. The middle ring, shaded in blue, displays the locations of all CentCs across the genome.  The inner ring, shaded in yellow, displays the locations of all CentC clusters greater than 20KB in length}
\label{circos}    
\end{figure}

@@

CentC copies in the maize genome exhibit tremendous diversity: the overall pairwise identity in our alignment was only 65\%, and $\sim$98\% of sites in the alignment had at least 2 variants.  Such diversity led us to ask whether genetic groups of CentC variants could be distinguished. We performed principle coordinate analyses (PCoA) from a genetic distance matrix estimated from our alignment, and assigned individual repeats to  genetic clusters following the approach of \citet{Patterson2006}.  We found 58 significant clusters, but observed no pattern of groupings that revealed chromosome specificity of CentC’s or the impact of historical tetraploidy (Figure \ref{pcoa}; Supplemental Table \ref{long}).

\begin{figure}[h]
\centering
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\includegraphics[width=1\textwidth]{Fig2_TWgroups}
\caption{Presence of CentC in each of the heirarchical groups.  The 58 clusters found to be statistically significant in forming genetic groups are represented on the x-axis and chromosome of origin on the y-axis. The size of each point is proportional to the log number of sequences in that group on that chromosome. CentC counts from chromosomes whose centromeres were derived from subgenome 1 are colored green and those from subgenome 2 are colored brown 
}
\label{pcoa}    
\end{figure}

@@

\begin{figure}[h]
\centering
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\includegraphics[width=1\textwidth]{Fig3_Heatmap}
\caption{CentC physical location and genetic relatedness for (a) chromosome 2 and (b) chromosome 5.  On the physical map, red lines show locations of numbered CentC clusters and blue blocks show the location of the active kinetichores.  Scale bar is in MB.  Below each physical map is shown a heatmap of genetic relatedness of each CentC to (top row) other copies within its island of tandem repeats delineated by dotted lines and (bottom row) all other copies on the chromosome.  Darker colors indicate higher relatedness.  The total number of CentC in each cluster is shown below the map
}
\label{heatmap}    
\end{figure}

The decreased genetic distance among CentCs in local clusters on chromosome 2 and 5 suggest that many of the genetic groupings discovered in our genome-wide analysis should correspond to local clusters of repeats. We see little evidence of this, however, as repeats within individual clusters are frequently found in different genetic groups as defined by principle coordinate analysis (Figure \ref{pcoa}).  A comparison of all pairs of CentC sequences reveals a likely explanation: of the $\sim 74$ million possible pairs,  approximately 6 million share $\geq 2$ mutations different from the genome-wide consensus, likely grouping in genetic clusters despite their physical distance.  Looking at several triplets at random from our alignment confirms that two sequences in one PCoA assignment sometimes share greater pairwise identity than two sequences adjacent to one another in different PCoA groups.  A simple forward simulation (see Methods) suggests this pattern could be due entirely to homoplasy rather than long-distance movement of CentC repeats.  By stochastically applying mutations to an initially homogeneous group of repeat sequences, we find that plausible parameter values produce $\sim 10$ million pairs of repeats sharing $\geq 2$ mutations.

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\centering
\includegraphics[width=1\textwidth]{Figure4_noedge}
\caption{(a) Mb of CentC in genomic libraries of maize and teosinte.  Box plots show data from \citet{Chia2012}.  Points show data for maize inbred B73 and the teosinte \emph{Z. luxurians} from \citet{Tenaillon2011}.  For comparison, the data point of maize inbred B73 in \citet{Chia2012} is shown with a tick mark on the box plot. (b) FISH images from an F1 hybrids between maize and the teosinte \emph{parviglumis}, and maize and the teosinte \emph{Z. luxurians}.  Upper row shows the karyotype for the teosinte and the lower row shows the maize karyotype.  Red=CentC, Green=CRM. 
}
\label{abundance}    
\end{figure}

@@

@@

%!CENT4 -- where to put you, cent4?
%We also sought to further characterize the chromosome specific repeat cent4 \citep{Page2001}.  Previous work showed that Cent4 probes lagged behind CentC probes in cell division \citep{Jiang2002}, suggesting that the repeat is not involved in the active kinetochore and rather located in the pericentromere.  A lack of enrichment of cent4 in ChIP data agrees with a pericentromeric location.  Using the published cent4 sequences, we identify the cent4 repeat as the LTR retrotransposon RLX\_sela in the transposable element database \citep{Baucom2009}.  Characterization of the TE consensus sequence shows that it contains repetitive motifs with homology to both knob and telomere repeats.  It lacks any of the protein sequences necessary for autonomous transposition, such as GAG and POL complexes.  Previous work in rice has documented the presence of nonautonomous LTR retrotransposons  in or near the centromere \citep{Jiang2002}.  However, the RLX\_sela sequence also appears to be missing some of the necessary primer binding sites that would distinguish it as a nonautonomous TE, suggesting that it may be a tandem repeat, and further sequencing of the region will be required to uncover the role of cent4.   Given that the centromere and pericentromere on chromosome 4 show signs of strong selection during domestication \citep{Hufford2012}, it is plausible that the increase in abundance of Cent4 seen in domesticated maize is due to the effects of linked selection on a locus important for domestication rather than novel insertions or duplications.

%!SUBGENOMES
Our study traces the changes in centromere repeat genetic relatedness across maize chromosomes to show how ancient tetraploidy and subsequent evolution has impacted centromeres.  Most interestingly, we show that most large arrays of CentC in the maize reference genome derive from maize1 subgenome, which is known to have lower gene loss and higher average expression than maize subgenome 2 \cite{Schnable2011}.  Regulatory microRNA’s are known to correspond to centromeric repeats \citep{ReinhartBartel2002}.  Therefore, it may be worthwhile to further investigate the potential correlation between centromere retention and gene expression as the phenomenon of higher expression of genes from one ancestor may be common across plants \citep{Cheng2012}.  There are several possible explanations for the differential contribution of subgenome 1 to the diversity of CentC in the extant maize  genome. We put forth several hypotheses that may explain the strong bias for CentC to be located within a block from subgenome 1. Subgenome 2 may have had severely reduced quantities of CentC and therefore never contributed equally.  The CentC from subgenome 2 may have decayed rapidly and thus less of it remains.  Lastly, the CentC from subgenome 1 may have expanded in copy number since the allopolyploidy event.  Though we are unable to conclusively identify the reason, the lack of a large number of identical tandem duplicates  and lack of close proximity highest genetic relatedness pairs suggests that any large scale expansion of CentC has not been recent.  Also, CentC’s in subgenome 2 cluster do not appear to have accumulated excess mutation, suggesting that they are not decaying more rapidly than their subgenome 1 counterparts.  We therefore believe that it is most likely that the subgenomes had an unequal contribution of CentC at the formation of the allopolyploid, though more research about the ancient parents is required.

%!RELATEDNESS
We also explored how genetic relatedness between CentC’s correlated with location along the physical map.  According to our PCoA analyses, CentC copies across the genome do not form distinct genetic groups correlating with past origin (Figure \ref{pcoa}).  The lack of subgenome or chromosome grouping in modern CentC suggests a little differentiation between the repeats since the ancestral divergence.  Given the lack of a genome wide pattern, we also wanted to investigate genetic relatedness on individual chromosomes.  We hypothesized that most copies of CentC arose from local duplication rather than transposition, and therefore the genetic distance between CentC’s would be lowest across CentC’s within a large tandem array.  \pb{This part is ugly LIKE YOUR GIT SKILLS, and needs help} From the reference genome, we chose to investigate chromosomes 2 and 5, since they have been sequenced from end to end, allowing for inferences about the relationship between diversity and physical proximity \citep{Wolfgruber2009}.  Furthermore, the role of the repeat arrays on chromosomes 2 and 5 appear very different, as the largest array on chromosome 2 interacts with the kinetochore, while no array on chromosome 5 does \citep{Wolfgruber2009}.  We find that repeats on chromosomes 2 and 5 are most highly genetically related to neighboring copies (Figure \ref{heatmap}).  This relationship was recapitulated in analyses using SpaGeDi, where CentC’s within approximately 10-50KB of one another were more genetically similar (Supplemental Data \ref{Supp_SpagediChr02} and \ref{Supp_SpagediChr05}).  Though we did not investigate the relationship between CentC location and genetic similarity on the other chromosomes due to incomplete sequencing of centromeres, we observe that many CentC’s within an array fall into the same  significant grouping in our Tracy-Widom clustering analysis, suggesting that high local similarity of CentC’s is a genome-wide phenomenon.  This local relatedness suggests that most CentC is evolving through tandem duplication and not a long distance mechanism such as transposition that had been previously suggested \citep{Shi2010}.  Concerning chromosomes 2 and 5, the higher local relatedness of all CentC clusters regardless of presence within the kinetochore suggests that CentC interaction with kinetochore proteins does not have a detectable change on their rates of evolution.

@@

%!DOMESTICATION
\jri{insert CITE, italics, probably refs to tables/figs as appropriate}  Previous cytogenetic work  identified  differences in centromere repeat content  between domesticated maize and its wild relatives Z. mays ssp. parviglumis,), Z. mays ssp. mexicana,, and Z. luxurians (hereafter parviglumis, mexicana, and luxurians) \citep{Albert2010} but was unable to quantify these differences. Our resequencing results show that while there is little difference in the distribution of CentC in tandem arrays, the total abundance of CentC has  decreased  \jri{can we say absolute abundance? we have genome size estimates, right?} during domestication, and we verify this with FISH in two maize-teosinte hybrid individuals (Figure \ref{abundance}). This result contrasts with the apparent increase in transposable element abundance  \citep{Chia2012}, suggesting that changes in CentC are not due to causes common to all repetitive sequence. \jri{ are we sure TEs have increase in absolute number? didn't chia et al. look at RPKM?  wouldn't that increase if overall genome size decreases? if we control for genome size, does TE content show and absolute increase?} The maize genome is smaller than its teosinte counterpart, largely due to differences in the abundance of heterochromatic knobs (Poggio et al 1998).  \citet{Zhang2012} have postulated an adaptive relationship between centromere size and genome size based on an observed correlation between centromere size and genome size across a number of grass species.  This model would propose that the decrease in CentC abundance seen post-domestication is due to selection for smaller active centromeres to complement the smaller overall genome size.  While our current data are insufficient to evaluate this conclusion, future work investigating differences in CentC content among maize landraces that vary in genome size (CITE) may provide an opportunity to further test this hypothesis.

@@

\begin{acknowledgements}
We wish to thank Pacific Biosciences for sequencing resources.  We thank Siddharth Bhadra-Lobo, Vince Buffalo, Anne Lorant, Gernot Presting, Lauren Sagara, Michelle Stitzer, and National Science Foundation summer exchange program interns Cesar Alvarez Mejia, Aurelio Hernandez Bautista for advice and helpful discussion. This project was funded by National Science Foundation grant IOS-0922703.
\end{acknowledgements}

@@

% For tables use
\begin{table}[]
\caption{ChIP Reads mapping to Unassembled from different Oat-Maize Addition (OMA) Lines.  Reads were mapped using Bowtie2 \citep{langmead2012} with the paramater -very sensitive-local.}\label{supp.gaby}       % Give a unique label

% For LaTeX tables use
\begin{tabular}{llll}
\hline\noalign{\smallskip}
File Key & Maize Chr & Percent Reads Aligning to Unassembled BACs\\
\noalign{\smallskip}\hline\noalign{\smallskip}				
JJ1BU (OMA 6.34) & 6 & 21.35\\
JJ1BR (OMA 1.36) & 1 & 15.56\\
JJ1CF (OMA 9.41) & 9 & 6.16\\
JJ1CH (OMA 8.05) & 8 & 6.69\\
JJ1CG (OMA 10.26) & 10 & 9.21\\
JJ1CI (OMA 8.05) & 8 & 8.5\\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}
